---
title: "Power calc"
author: "Dan Weinberger"
date: "March 11, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lme4)
library(lubridate)
library(MASS)
library(RCurl)
library(viridis)
library(grDevices)
library(pbapply)
```

## Read in and format data
This is state-level data from Brazil, In this instance, we are reading directly from Github

```{r cars, fig.width=8, fig.height=6}
x <- getURL("https://raw.githubusercontent.com/weinbergerlab/Brazil_state/master/prelog_Brazil_state_processed_data.csv")
ds1<-read.csv(text=x)
ds1<-ds1[,c('age_group', 'date','J12_18', 'ach_noj')] #Keep only variables we need
ds1<-ds1[substr(ds1$age_group,1,2) %in% c('09','08'),] #age group 9 (<12m olds) only
ds1$age_group<-factor(ds1$age_group)
ds2<-split(ds1, ds1$age_group)
time_points1<-as.Date(unique(ds1$date))

source("power_calc_custom_functions.R")

par(mfcol=c(2,2), mar=c(3,4,2,2))
plot(as.Date(ds2[['09_3_35']]$date) , ds2[['09_3_35']]$J12_18, bty='l', type='l', ylab='Pneumonia hospitalizations',xlab='date', ylim=c(0, max( ds2[['09_3_35']]$J12_18)), main='Large State, <12m')
abline(v=as.Date('2010-01-01'), col='gray', lty=2)

plot(as.Date(ds2[['08_3_35']]$date) , ds2[['08_3_35']]$J12_18, bty='l', type='l', ylab='Pneumonia hospitalizations',xlab='date', ylim=c(0, max( ds2[['08_3_35']]$J12_18)), main='Large State, 80+y')
abline(v=as.Date('2010-01-01'), col='gray', lty=2)

plot(as.Date(ds2[['09_1_12']]$date),ds2[['09_1_12']]$J12_18, bty='l', type='l', ylab='Pneumonia hospitalizations',xlab='date',ylim=c(0, max( ds2[['09_1_12']]$J12_18)) ,main='Small State, <12m')
abline(v=as.Date('2010-01-01'), col='gray', lty=2)

plot(as.Date(ds2[['08_1_12']]$date),ds2[['08_1_12']]$J12_18, bty='l', type='l', ylab='Pneumonia hospitalizations',xlab='date',ylim=c(0, max( ds2[['08_1_12']]$J12_18)) ,main='Small State, 80+y')
abline(v=as.Date('2010-01-01'), col='gray', lty=2)

```

Ave cases
```{r}
ave.pre<-sapply(ds2, function(x) {
  x$date<-as.Date(x$date)
  ave.pre<-mean(x$J12_18[x$date<=as.Date("2010-02-01")])
  
}
  )
ave.pre.kids<-ave.pre[substr(names(ave.pre),1,2)=='09'& !grepl('A',names(ave.pre))]
range(ave.pre.kids)

ave.pre.elderly<-ave.pre[substr(names(ave.pre),1,2)=='08' & !grepl('A',names(ave.pre))]
range(ave.pre.elderly)
```


## Set parameters

```{r parms, echo=FALSE}
set.nsim=500  #How many simulations per stratum?
set.post.start.date<- '2010-01-01'  #When was vaccine introduced?


```

## Simulate post-vaccine data

```{r, eval=F}
#######################
set.seed(123)
ve<- c(0.5, 0.6, 0.7, 0.8, 0.9) #IRR estimates to simulate
sim.ds.ve<-vector("list", length(ve)) 
for( i in 1:length(ve)){
sim.ds.ve[[i]]<-lapply(ds2,ts.extract.func, outcome.name='J12_18',
                       covar.names=c('ach_noj'),
                    nsim=set.nsim, #Number of simulations, as set above
                    ve.irr=ve[i],  #True vaccine effect
                    post.start=set.post.start.date #Date of vaccine introduction
               )
}
names(sim.ds.ve)<-ve
re.sd<- sapply(sim.ds.ve[[1]], '[[', 're.sd' )
saveRDS(re.sd, 'brazil.state.phi.sd.rds')
saveRDS(sim.ds.ve, 'sim.ds.ve.rds')

```

```{r}
pre.dates<-1:(which( unique(ds1$date) ==set.post.start.date )-1)
pre.vax.ave<-sapply(ds2, function(x) mean(x$J12_18[pre.dates]))
```

```{r}
re.sd<-readRDS('brazil.state.phi.sd.rds')
sim.ds.ve <- readRDS('sim.ds.ve.rds')
```

check
```{r}
nat.sim<-sim.ds.ve[[1]][['09_A_AA']]
matplot(nat.sim$preds.stage2, type='l')

jan<-nat.sim$preds.stage2[seq(from=1,to=144,by=12),]
rowMeans(jan)
matplot(jan, type='l')

```


#Plot some of the simulated data (10 draws each)

```{r, fig.width=8, fig.height=6}
plot.cols<-c('#a6cee3',
'#1f78b4',
'#b2df8a',
'#33a02c',
'#fb9a99',
'#e31a1c',
'#fdbf6f',
'#ff7f00',
'#cab2d6',
'#6a3d9a')
set.seed(123)
plot.cols<-sample(plot.cols)
sapply(sim.ds.ve[[1]] ,function(x)  mean(x$preds.stage2))
n.states.plot<-5

par(mfcol=c(2,2), mar=c(3,4,2,2))

matplot(sim.ds.ve[[1]][['09_1_12']]$preds.stage2[,1:n.states.plot], type='l', bty='l',lty=1, ylab='N cases', main='Small state, 50% decline', col=plot.cols[1:n.states.plot])
abline(v=pre.dates[length(pre.dates)], lty=2, col='gray')
matplot(sim.ds.ve[[1]][['09_3_35']]$preds.stage2[,1:n.states.plot], type='l',lty=1, bty='l',ylab='N cases',
        main='Large state, 50% decline',col=plot.cols[1:n.states.plot])
abline(v=pre.dates[length(pre.dates)], lty=2, col='gray')

matplot(sim.ds.ve[[5]][['09_1_12']]$preds.stage2[,1:n.states.plot], type='l', bty='l',lty=1, ylab='N cases',main='Small state, 10% decline',col=plot.cols[1:n.states.plot])
abline(v=pre.dates[length(pre.dates)], lty=2, col='gray')
matplot(sim.ds.ve[[5]][['09_3_35']]$preds.stage2[,1:n.states.plot], type='l',lty=1, bty='l',ylab='N cases',main='Large state, 10% decline',col=plot.cols[1:n.states.plot])
abline(v=pre.dates[length(pre.dates)], lty=2, col='gray')

#80plus years

matplot(sim.ds.ve[[1]][['08_1_12']]$preds.stage2[,1:n.states.plot], type='l', bty='l',lty=1, ylab='N cases', main='Small state, 50% decline', col=plot.cols[1:n.states.plot], ylim=c(0,max(sim.ds.ve[[5]][['08_1_12']]$preds.stage2[,1:n.states.plot])) )
abline(v=pre.dates[length(pre.dates)], lty=2, col='gray')

matplot(sim.ds.ve[[1]][['08_3_35']]$preds.stage2[,1:n.states.plot], type='l',lty=1, bty='l',ylab='N cases',
        main='Large state, 50% decline',col=plot.cols[1:n.states.plot], ylim=c(0,max(sim.ds.ve[[5]][['08_3_35']]$preds.stage2[,1:n.states.plot])) )
abline(v=pre.dates[length(pre.dates)], lty=2, col='gray')

matplot(sim.ds.ve[[5]][['08_1_12']]$preds.stage2[,1:n.states.plot], type='l', bty='l',lty=1, ylab='N cases',main='Small state, 10% decline',col=plot.cols[1:n.states.plot], ylim=c(0,max(sim.ds.ve[[5]][['08_1_12']]$preds.stage2[,1:n.states.plot])) )
abline(v=pre.dates[length(pre.dates)], lty=2, col='gray')

matplot(sim.ds.ve[[5]][['08_3_35']]$preds.stage2[,1:n.states.plot], type='l',lty=1, bty='l',ylab='N cases',main='Large state, 10% decline',col=plot.cols[1:n.states.plot],  ylim=c(0,max(sim.ds.ve[[5]][['08_3_35']]$preds.stage2[,1:n.states.plot])) )
abline(v=pre.dates[length(pre.dates)], lty=2, col='gray')

```

## Run ITS on the simulated data
(note set to eval=F)
```{r,eval=F}
rr.ests.month.irr <-vector("list", length(sim.ds.ve)) 

# Start the clock!
ptm <- proc.time()
for(i in 1: length(sim.ds.ve)){
rr.ests.month.irr[[i]]<-pbsapply(sim.ds.ve[[i]], est.vax.eff,time_points.select=time_points1,  overdisperse1=T ,n.season=12 , simplify='array'   )
} 
# Stop the clock
proc.time() - ptm
rr.ests.compile<-c(rr.ests.month.irr )

saveRDS(rr.ests.compile, 'sim1.rds')
```

```{r}
rr.ests.compile<-readRDS('sim1.rds')
```

## What proportion of the simulations had a RR<1

```{r}
plot(rr.ests.compile[[1]][,'50%','09_1_12'], ylim=c(0.2,1.5), bty='l')
arrows(y0=rr.ests.compile[[1]][,'2.5%','09_1_12'],y1=rr.ests.compile[[1]][,'97.5%','09_1_12'],x0=1:length(rr.ests.compile[[1]][,'50%','09_1_12']), length=0)
points(rr.ests.compile[[1]][,'50%','09_1_12'], pch=16, col=rgb(0,0,0,0.6))
abline(h=0.5, lty=3, col='gray')
abline(h=1, lty=2, col='red')

plot(rr.ests.compile[[4]][,'50%','09_1_12'], ylim=c(0.2,1.5), bty='l')
arrows(y0=rr.ests.compile[[4]][,'2.5%','09_1_12'],y1=rr.ests.compile[[4]][,'97.5%','09_1_12'],x0=1:length(rr.ests.compile[[4]][,'50%','09_1_12']), length=0, col='gray')
points(rr.ests.compile[[4]][,'50%','09_1_12'], pch=16, col=rgb(0,0,0,0.6))
abline(h=0.8, lty=3, col='gray')
abline(h=1, lty=2, col='red')
```



## Calculate Power
```{r}
power.month<-lapply(rr.ests.compile, function(x){
  y<-x[,"97.5%",]  
  power<-apply(y, 2, function(z) mean(z<1))
  return(power)
})
```


## Calculate bias
```{r}
mse.func<-function(x, true.rr){
  y<-x[,"50%",]  
  mse<-apply(y, 2, function(z) mean( (z -true.rr)^2  ))
  return(mse)
}

mse<-mapply(mse.func, rr.ests.compile, ve, SIMPLIFY=F)
```

## Calculate coverage of the CIs
```{r}
cov.func<-function(x, true.rr){
  cov<-apply(x, 3, function(z){
     mean( (true.rr> z[,'2.5%'] &  true.rr< z[,'97.5%']  )   ) 
  })
    return(cov)
}
ci.cov<-mapply(cov.func, rr.ests.compile, ve, SIMPLIFY=F)

```




## Unexplained variability in pre-vaccine period vs power and MSE
```{r, fig.width=10, fig.height=2.25}
par(mfrow=c(1,length(power.month)), mar=c(5,4,3,1))
grp.labels<-names(power.month[[1]])
age.grp<-as.numeric(as.factor(substr(grp.labels,1,2)))
cols.grp<-c('darkgray', 'black')
pch.grp=c(16,17)

for(i in 1:length(power.month)){
plot(re.sd, power.month[[i]], bty='l', col=cols.grp[age.grp],pch=pch.grp[age.grp], bty='l',  xlab='Unexplained variability (SD)', ylim=c(0,1),ylab='Power',main=paste0('Expected IRR ',ve[i]))
}

# for(i in 1:length(mse)){
# #plot(re.sd, power.month[[i]], bty='l', main='Random noise vs power')
# plot(re.sd, sqrt(mse[[i]]), bty='l', xlab='Unexplained variability (SD)',ylab='Bias (sqrt(MSE))', main=paste0('Expected IRR ',ve[i]), ylim=c(0,0.2), col=cols.grp[age.grp], pch=pch.grp[age.grp])
# #axis(side=1, at=c(2:10),labels=round(exp(2:10),-1) )
# }

```

## Number of cases vs power and bias
```{r, fig.width=10, fig.height=2.25}
par(mfrow=c(1,length(power.month)), mar=c(5,4,3,1))
grp.labels<-names(power.month[[i]])
age.grp<-as.numeric(as.factor(substr(grp.labels,1,2)))
cols.grp<-c('darkgray', 'black')
pch.grp=c(16,17)

for(i in 1:length(power.month)){
plot(log(pre.vax.ave), power.month[[i]], bty='l',xaxt='n', col=cols.grp[age.grp],pch=pch.grp[age.grp], bty='l',  xlab='Ave. cases per month', ylim=c(0,1),ylab='Power',main=paste0('Expected IRR ',ve[i]))
#plot(log(pre.vax.ave), power.month[[i]], bty='l', xaxt='n', xlab='Ave cases per month', main=paste0('Expected IRR ',ve[i]), ylim=c(0,1), col=cols.grp[age.grp])
axis(side=1, at=c(2:10),labels=round(exp(2:10),-1) )
}

# for(i in 1:length(mse)){
# #plot(re.sd, power.month[[i]], bty='l', main='Random noise vs power')
# plot(log(pre.vax.ave), sqrt(mse[[i]]), bty='l', xaxt='n', xlab='Ave. cases per month',ylab='Bias (sqrt(MSE))', main=paste0('Expected IRR ',ve[i]), ylim=c(0,0.2), col=cols.grp[age.grp], pch=pch.grp[age.grp])
# axis(side=1, at=c(2:10),labels=round(exp(2:10),-1) )
# }

```


## Unexplained variability in pre-vaccine observed data vs MSE estimates
```{r,fig.width=10, fig.height=2}
# par(mfcol=c(1,length(power.month)), mar=c(4,4,1,1))
# for(i in 1:length(power.month)){
# #plot(re.sd, power.month[[i]], bty='l', main='Random noise vs power')
# plot(re.sd, sqrt(mse[[i]]), bty='l', xaxt='n', xlab='Unexplained variabiity in observed data', main=paste0('Expected IRR ',ve[i]), ylim=sqrt(c(0,0.03)), col=cols.grp[age.grp], pch=pch.grp[age.grp])
# axis(side=1, at=c(2:10),labels=round(exp(2:10),-1) )
# }

```

## Number of cases vs unexplained variability

```{r, fig.width=5, fig.height=5}
cols.grp<-c('darkgray', 'black')
pch.grp=c(16,17)
par(mfrow=c(1,1))
plot(log(pre.vax.ave), re.sd,col=cols.grp[age.grp], pch=pch.grp[age.grp], bty='l', ylab="Unexplained variability", xlab="log(N monthly cases)")
```


##How many years do you need?
Focus on datasets with an IRR of 0.8
```{r}
rr.0.8<-sim.ds.ve[['0.8']]
trim.function<-function(ds, trim.length){
  ds$preds.stage2 <- ds$preds.stage2[-c(1:trim.length),]
  ds$covars<-ds$covars[-c(1:trim.length),,drop=F]
  return(ds)
}
trim.12<-lapply(rr.0.8, function(x) trim.function(ds=x, trim.length=12) )
trim.24<-lapply(rr.0.8, function(x) trim.function(ds=x, trim.length=24) )
trim.36<-lapply(rr.0.8, function(x) trim.function(ds=x, trim.length=36) )
trim.48<-lapply(rr.0.8, function(x) trim.function(ds=x, trim.length=48) )
trim.60<-lapply(rr.0.8, function(x) trim.function(ds=x, trim.length=60) )
trim.72<-lapply(rr.0.8, function(x) trim.function(ds=x, trim.length=72) )

trim.list<-list(trim.12,trim.24,trim.36, trim.48, trim.60, trim.72)
trim.lengths<-c(12,24,36,48,60,72)
time_points.trim<-lapply(trim.lengths, function(x)  time_points.trim<-time_points1[-c(1:x)] )
```


```{r, eval=F}
#Initialize list
rr.ests.month.irr.trim <- vector("list", length(trim.list)) 

# Start the clock!
ptm <- proc.time()
for(i in 1: length(trim.list)){
rr.ests.month.irr.trim[[i]]<-pbsapply(trim.list[[i]], est.vax.eff, time_points.select=time_points.trim[[i]], overdisperse1=T ,n.season=12 , simplify='array'   )
} 
# Stop the clock
proc.time() - ptm

rr.ests.compile.trim<-c(rr.ests.month.irr.trim )
saveRDS(rr.ests.compile.trim, 'sim1.trim.rds')
```

```{r}
rr.ests.compile.trim<-readRDS('sim1.trim.rds')
```


Power

```{r}
power.month.trim<-lapply(rr.ests.compile.trim, function(x){
  y<-x[,"97.5%",]  
  power<-apply(y, 2, function(z) mean(z<1))
  return(power)
})
power.month.trim.array<-do.call('cbind', power.month.trim)
```

#Calculate bias
```{r}

mse.trim<-mapply(mse.func, rr.ests.compile.trim, true.rr=0.8, SIMPLIFY=F)
```

```{r, fig.width=6, fig.height=6}
pal1<-viridis_pal(option='cividis')(length(trim.lengths))
legend_image <- as.raster(matrix(pal1, ncol=1))
par(mfrow=c(1,1))
par( mar=c(4,4,1,1))
for(i in 1:length(power.month.trim)){
#test.range <-c(4:6)
#for(i in test.range){
if(i==1){
plot(re.sd, power.month.trim[[i]], bty='l', xaxt='n', xlab='Unexplained variability (SD)',ylab='Power', ylim=c(0,1), col=pal1[i], pch=16)
axis(side=1, at=c(2:10),labels=round(exp(2:10),-1) )
}else{
  points(re.sd, power.month.trim[[i]], col=pal1[i], pch=16)

}
}
rasterImage(legend_image, 0.3, 0.6, 0.31,0.9)
text( x=rep(0.32, 3) , y=seq(from=0.9, to=0.6, length.out=3) , rev(seq(from=12, to=72 ,by=24)))
text(x=0.3, y=0.94, 'Baseline Months (N)', adj=c(0,0), xpd=NA)

# for(i in 1:length(mse.trim)){
# if(i==1){
# plot(re.sd, sqrt(mse.trim[[i]]), bty='l', xaxt='n', xlab='Unexplained variability (SD)',ylab='Bias (RMSE)', ylim=c(0,0.26), col=pal1[i], pch=16)
# axis(side=1, at=c(2:10),labels=round(exp(2:10),-1) )
# }else{
#   points(re.sd, sqrt(mse.trim[[i]]), col=pal1[i], pch=16)
# 
# }
# }
# #rasterImage(legend_image, 0.075, 0.16, 0.085,0.24)
#text( x=rep(0.095, 3) , y=seq(from=0.24, to=0.16, length.out=3) , rev(seq(from=12, to=72 ,by=24)))
#text(x=0.075, y=0.26, 'Baseline Months (N)', adj=c(0,0), xpd=NA)
```

```{r}
pal1<-viridis_pal(option='cividis')(length(re.sd))
sd.rank<-rank(re.sd+rnorm(length(re.sd),0,0.01))

power.month.mat<-matrix(unlist(power.month), nrow=length(power.month[[1]]))
matplot((84-trim.lengths),100*t(power.month.mat), type='l', bty='l', main='Power vs number of pre-vaccine months', xlab='Number of pre-vaccine months', ylab='Power', ylim=c(0,100), col=pal1[sd.rank])
```
Small states (<100 cases/month) didn't benefit much from increasing the number of pre-vaccine months--their power is still weak.

This shows for states that start with low power (light gray), the absoute improvement with adding more years is smaller (which makes sense); but the relative improvement in power is similar in states with strong and weak power to start
```{r}
#Power improvements 
improve1<- power.month.mat[,1]-power.month.mat[,6]#difference in power
improve2<- power.month.mat[,1]/power.month.mat[,6]#ratio of power
col1<- rgb(0,0,0, alpha=power.month.mat[,1]) #darker color indicates stronger power in column 1

plot(log(pre.vax.ave), improve1, bty='l', pch=16, main='Improvement of power with more years vs number of cases')

pal2<-viridis_pal(option='cividis')(length(re.sd))

init.power.rank<-rank(power.month.mat[,1]+rnorm(length(re.sd),0,0.01))
par(mfcol=c(1,1))
plot(re.sd, improve1, bty='l', pch=16, main='Improvement of power with more years vs unexplained random noise in the pre period', col=pal2[init.power.rank])

plot(re.sd, improve2, bty='l', col=pal2[init.power.rank], pch=16, main='Improvement of power with more years vs unexplained random noise in the pre period')
```

